---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Ayden Schirmacher"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
toc: TRUE
---

\listoffigures

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2, warning=FALSE, message=FALSE}
remove(list=ls())

#1
getwd() #get working directory

library(tidyverse) 
library(agricolae)
library(dplyr)
library(here)
library(ggplot2)
library(corrplot)
library(ggthemes)
library(crayon)
#^^loading packages

raw_data<-read.csv( #reading in data
  file=here("Data","Raw","NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
  stringsAsFactors=TRUE)%>%
  mutate(s_date=mdy(sampledate)) #setting date column as date

#2
theme_A07 <- theme_base() + 
  theme(
    text = element_text(color='black', size=10, face = 'italic'),
    panel.grid.minor = element_line(color="gray87"),
    panel.grid.major = element_line(color="gray87"),
    plot.background = element_rect(color='black', fill='snow1'),
    axis.ticks = element_line(size=0.25))

theme_set(theme_A07) #creating theme for outputs

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:

>H0: The mean lake temperature recorded during July does not change with depth across all lakes

>Ha: The mean lake temperature recorded during July does change with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot, message=FALSE, warning=FALSE, fig.cap='Temperature by Depth Across All Lakes'}

#4
processed_data <- raw_data %>%
  filter(month(s_date) == 7) %>% #filtering for july
  select(lakename, year4, daynum, depth, temperature_C) %>% #selecting variables
  filter(!is.na(lakename), !is.na(year4), !is.na(daynum), 
         !is.na(depth), !is.na(temperature_C)) #filtering out na values

#5
ggplot(processed_data, aes(x = depth, y = temperature_C))+ #setting x and y
  geom_point(size = 2, alpha = 0.25, color = "steelblue")+
  #^^ point information
  geom_smooth(method = "lm",linewidth = 0.65, col = "red3")+
  #^^ trend line
  geom_text(x=12.5,y=20, #where the label goes on graph
    label= "temperature = -1.94621*depth + 21.95597 ", size=3)+
  #^^labeling the expression/equation on graph
  ylim(0,35)+ #setting y limits
  labs(x = "depth (m)",y = "temperature (C)")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The relationship is negative, indicating that an increase in temperature corresponds with an decrease in depth, and a decrese in temperature corresponds with an increase in depth. The distribution of points is fairly even around the line of best fit, though there are some points that do not apply to this trend as the depth increases. This suggests that there is a fairly normal distribution of points with a slight right skew. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression, message=FALSE, warning=FALSE}
#7
lin_reg <- lm(data = processed_data, temperature_C ~ depth)
summary(lin_reg) 
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: With 9726 degrees of freedom and a p-value of <0.01, there is convincing evidence to reject the null hypothesis, and say that the mean lake temperature recorded during July does change with depth across all lakes. This is confirmed by the model's output that a one unit change in depth (meters) results in a -1.94621 unit change in temperature (C) across all lakes and an R-squared value that indicates that depth explains 73.87% of the variability in temperature across all lakes. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model, message=FALSE, warning=FALSE}
#9
multi_reg_aic<-lm(data = processed_data, temperature_C ~ year4 + daynum + depth)
step(multi_reg_aic)
```

```{r temperature.model optimized, message=FALSE, warning=FALSE}
#10
multi_optimized<-lm(data=processed_data, temperature_C ~ daynum + year4 + depth)
summary(multi_optimized)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables includes all of the variables originally selected -- daynum, year4, and depth. The model explains 74.12% of the variance in temperature, which is an improvement, although a small one, over the model which only used depth. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model, warning=FALSE, message=FALSE}
#12
lakename_anova<-aov(data=processed_data, temperature_C ~ lakename)
summary(lakename_anova)
```

```{r anova.model.as.lm, warning=FALSE, message=FALSE}
lakename_lm<-lm(data=processed_data, temperature_C ~ lakename)
summary(lakename_lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes. For the ANOVA test, given that the F-value represents the the ratio of the between-group variance to the within-group variance, having an F-value of 5- against 8 degrees of freedom on a p-alues <0.01 suggests convincing evidence to reject the null hypothesis that the means across lakes are the same. In the linear model the p-values for all coefficients for all lakenames are <0.01, indicating convincing evidence of difference between all lakes and the baseline lake of Central Lake.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, warning=FALSE, message=FALSE, fig.cap='Temperature by Depth Across All Lakes'}
#14.
ggplot(processed_data, aes(x = depth, y = temperature_C, col = lakename))+
  geom_point(alpha = 0.35)+
  ylim(0,35)+
  geom_smooth(method = "lm", se = FALSE, linewidth=0.85)+
  labs(x="depth (m)", y="temperature (C)")
```

```{r scatterplot.3, warning=FALSE, message=FALSE, fig.cap='Temperature by Depth Across Individual Lakes'}
#14. 
ggplot(processed_data, aes(x = depth, y = temperature_C, col = lakename))+
  geom_point(alpha = 0.35)+
  ylim(0,35)+
  geom_smooth(method = "lm", se = FALSE, linewidth=0.85, col="red")+
  labs(x="depth (m)", y="temperature (C)", title = "Temperature by depth across individual lakes")+
  facet_wrap(~ lakename) 
#^^ making a plot for each lake
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test, warning=FALSE, message=FALSE}
#15
TukeyHSD(lakename_anova)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake (p-value ~ 0.2241586) and Ward Lake (p-value ~0.7827037). With p-values larger than 0.10, we do not have sufficient evidence to reject the null hypothesis of no difference in means. None of the lakes have a mean temperature distinct from ALL other lakes. 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A two-sample t-test. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test, message=FALSE, warning=FALSE}
july_only2lakes<- processed_data%>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

two_sample<-t.test(temperature_C ~ lakename, data=july_only2lakes)
two_sample
  
```

>Answer: The test returns that we do not have sufficient evidence to reject the null hypothesis that the true difference in means is zero between Crampton and Ward Lakes. Furthemore, since the confidence interval includes zero, we cannot conclude a significant difference between the temperatures of Crampton Lake and Ward Lake, supporting the p-value result. A t-score of 1.1181 is relatively small, indicating that the difference between the two lake temperatures is minor compared to the variability in the data. The Tukey's test from #16 estimated a mean difference between the lakes of -0.8932661, with a non-significant p-value of 0.97, leading to insufficient evidence for rejecting the null. This is the same result as the two-sample t-test. 


