---
title: "Assignment 3: Data Exploration"
author: "Ayden Schirmacher"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  read files, message = FALSE}
#loading packages (already in system, now just making sure they're on)
install.packages("tidyverse")
install.packages("lubridate")
install.packages("here")
library(tidyverse)
library(lubridate)
library(here)

#reading neonics and litter files, "here" is set to the ENV repository, specify that we are looking in Data\Raw folder
neonics<-read.csv(
  file = here('Data','Raw','ECOTOX_Neonicotinoids_Insects_raw.csv'),
  stringsAsFactors = T)

litter<-read.csv(
  file = here('Data', 'Raw', 'NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
  stringsAsFactors = T)

```
 
## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: the ecotoxicology of neonicotinoids is important because, as pesticides, they impact a whole host of ecosystems. Knowing the ecotox of the product helps to understand what species the product might target and how this will impact pollinator and pest systems. Furthermore, we can understand better how the runoff from pesticide use will impact surrounding environments and resources, as well as how plants and animals may become resistant to the product.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter and woody debris play a large role in soil health and bottom canopy ecosystems, especially in forest ecosystems. Understanding this process will help to understand nurtrient cycling and soil chemistry, and how these impact fungal and microbio communities. There are also large implications for carbon sequestration as the organic matter decomposes into the ground and for forest management (i.e. fire suppression, forest health, etc.)

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. According to the manual, "Mass data for each collection event are measured separately for the following functional groups, to an accuracy of 0.01 grams. Weights < 0.01g are reported and may indicate presence of a given functional group, identified in the sorting process, but not present at detectable masses" (p.3)
 2. Spatial sampling design varies based on height of vegetation and percent of vegetation that is classified as "woody". 
 3. Temporal sampling design varies between sites classified as deciduous vs. evergreen. More frequent sampling was conducted at deciduous sites. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

> Answer: There are 4,623 observations of 30 variables

```{r data dimensions}
str(neonics) #shows number of rows and colums in dataset using dplyr package
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r summary neonics effects}
sort(summary(neonics$Effect)) #sort the number of observations in the neonics effects column
```

> Answer: The most common effects studied in the neonics dataset were population with 1,803 observations; mortality with 1,493 observations; and behavior with 360 observations. These are likely the most commonly studied because they are important baselines for effects studies -- you cannot know true effects without knowing the overall population health and stability. Furthermore, it is likely that these are easier to study than say, intoxication. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

> Answer: The most common observation was "other". However, the six most common specific species are 1. Honey Bee, 2. Parasitic Wasp; 3. Buff Tailed Bumblebee; 4. Carniolan Honey Bee; 5. Bumble Bee; and 6. Italian Honeybee. All of these are types of bees or wasps belonging to the Hymenoptera order, which includes both bees and wasps. All play crucial roles in pollination. Because of this role in pollination, knowing how pesticides and insecticides impact these organisms is very important -- without them, breeding and producing agricultural products would be much, much harder. 

```{r summary neonics common name}
sort(summary(neonics$Species.Common.Name)) #sort the number of observations in the neonics 'species common name' column
```

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

> Answer: The 'Conc.1.Author' column is a factor class. This is because the column contains some non-numeric values (i.e. "NR/"), so the R software automatically reads the column as non-numeric in order to make sense of it. 

```{r finding class of conc.1.author for neonics}
class(neonics$Conc.1..Author.) #find class of column
print(neonics)
```


## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r frequency plot in neonics}
ggplot(neonics, aes(x = Publication.Year)) + #establish publication year (time) as independent variable
  geom_freqpoly(binwidth = 1) + #frequency plot code
  labs(title = "Number of Studies Conducted by Publication\n Year in Neonics Dataset, 1980-2025", 
       x = "Year", 
       y = "Number of Studies") +
  theme_minimal() #make professional
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r color frequency plot in neonics}
ggplot(neonics, aes(x = Publication.Year, color = as.factor(Test.Location))) + #color the graph by location
  geom_freqpoly(binwidth = 1) + #size of lines
  labs(title = "Number of Studies Conducted by Publication\n Year in Neonics Dataset, 1980-2025", 
       x = "Year", 
       y = "Number of Studies") +
  theme_minimal()

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The graph shows that the number of studies was largest between 2010 and 2015, with the larges spike being in ~2014 and a very steep dropoff after that year. By using color differentiation, we can see that the most common test location is in natural fields between ~1992 and ~2000, before being overtaken by testing locations in labs between ~2000 and ~2020. However, there was a brief period from ~2008 to ~2010 where natural field sampling was the predominant location. Articficial fields and undeterminable fields present very small proportions of the overall test location data. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r boxplot for endpoint in neonics}
ggplot(neonics, aes(x=Endpoint))+ #endpoint as independent variable 
  geom_bar(width=0.8, fill="maroon", color="black")+ #fill bars with color to make pretty
         labs(x="Endpoints",
              y="Count",
              title="Endpoint observation count\n for Neonics dataset")+
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) #rotate and align x-axis labels

```

> Answer: The two most common endpoints are LOEL and NOEL. According to the ECOTOX_CodeAppendix, these represent the information listed below. Both are a part of the Terrestrial database usage (p. 721-722)
- LOEL: Lowest-observable-effect-level: lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls (LOEAL/LOEC)
- NOEL: No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test (NOEAL/NOEC)

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

> Answer: in August 2018, data was collected on August 2nd and August 30

```{r make Date class in litter}
class(litter$collectDate) #check class of collectDate, its a factor
litter$collectDate<-as.Date(litter$collectDate) #change to Date class
class(litter$collectDate) #confirm change

unique(litter$collectDate, 2018-08) #check dates in august 2018 that data was collected
```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

>Answer: The information obtained from unique counts only how many different plot IDs there are and tells what the names of those unique plots are. Summary expands on this information by telling how many times a study was taken from those unique sites. 

```{r unique v. summary in litter}
unique(litter$plotID) #using unique function
summary(litter$plotID) #using summary function for more detail
```

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r bar graph for types of litter}
ggplot(litter, aes(x=functionalGroup))+ #types of litter as independent variable
  geom_bar(width=0.7, fill="steelblue", color="black")+ #color for bins
  labs(x="Types of Litter",
       y="Count",
       title="Types of litter collected at Niwot Ridge sites")+
  theme(axis.text.x = element_text(angle = 45, hjust=1)) #rotate and align x-axis labels

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r boxplot dryMass/functionalgroup}
ggplot(litter, aes(x=functionalGroup, y=dryMass))+ #use dryMass as dependent variable
  geom_boxplot(width=0.7)+ #make boxplot
  labs(x="Types of Litter",
       y="Mass of Dry Material",
       title="Mass of litter organized \nby type collected from Niwot Ridge")+
  theme(axis.text.x = element_text(angle = 45, hjust=1)) #adjust titles for legibility
```

```{r violin plot dryMass/functionalgroup}
ggplot(litter) +  
  geom_violin(aes(x = functionalGroup, y = dryMass), draw_quantiles=c(0.25, 0.5, 0.75))+ #use dryMass as dependent variable
  labs(x = "Types of Litter",
       y = "Mass of Dry Material",
       title = "Mass of Litter Organized \nby Type Collected from Niwot Ridge")+
    theme(axis.text.x = element_text(angle = 45, hjust=1)) #adjust titles for legibility
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The violin plot is not showing any useful information. This could be due to the fact that the distribution of litter types was fairly evenly distributed, so the density of observations is not extremely variable. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles is the highest, followed by mixed with the most consistent high biomass, though twigs/branches has a high outlier. 
