---
title: "Assignment 5: Data Visualization"
author: "Ayden Schirmacher"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
remove(list=ls())
#1 loading packages to library and check workspace
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
getwd()

#set variable name for workspace
processed = "./Data/Processed_KEY"

#read in csv files
lakes_data<-read_csv(
  here(processed,"NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"))

niwot_ridge<-read_csv(
  here(processed,"NEON_NIWO_Litter_mass_trap_Processed.csv"))

#2 check that dates are read as dates
class(lakes_data$sampledate)
class(niwot_ridge$collectDate)
```

## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
custom_theme <- theme_base() + 
  theme(
    text = element_text(color='black', size=10, face = 'italic'),
    panel.grid.minor = element_line(color="gray87"),
    panel.grid.major = element_line(color="gray87"),
    plot.background = element_rect(color='black', fill='snow2'),
    axis.ticks = element_line(size=0.25))

theme_set(custom_theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
lakes_data%>%
ggplot(aes(x=tp_ug, y=po4, color=lakename))+
scale_color_manual(values = c("Peter Lake" = "navy", 
                               "Paul Lake" = "darkorange2"))+
#^^ manual selecting of colors for each lake
  geom_point(alpha=0.5)+
  ylim(0,45)+
  geom_smooth(method = lm, se=FALSE, size=0.75)+
  labs(x="total phosphorus",
      y="total phosphate",
      title="Ratio of phosphorus to
        phosphate in the \nPeter and Paul Lakes")
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 creating three boxplots
temp_plot<-ggplot(lakes_data, aes(x=month, y=temperature_C, color=lakename))+
  #^^ selecting variables and coloring by lake name
scale_color_manual(values = c("Peter Lake" = "navy", 
                               "Paul Lake" = "darkorange2"))+
  geom_boxplot(aes(x=factor(month, levels=1:12, labels=month.abb)))+
  labs(x=element_blank(), y="temperature (°C)", #eliminating x labels
       title="Monthly temperature of water in Peter and Paul Lakes")+
    theme(legend.position = "none") #eliminating legend

tp_plot<-ggplot(lakes_data, aes(x=month, y=tp_ug, color=lakename))+
scale_color_manual(values = c("Peter Lake" = "navy", 
                               "Paul Lake" = "darkorange2"))+
  geom_boxplot(aes(x=factor(month, levels=1:12, labels=month.abb)))+
    #^^ making all months visible to appropriately visualize annual data
  labs(x=element_blank(), y="total P (µg)", #eliminating x labels
       title="Monthly P levels in water of Peter and Paul Lakes")+
  theme(legend.position = c(0.87,0.5)) 
  #^^ placing legend inside of the middle plot so that the plots stay aligned
  
tn_plot<-ggplot(lakes_data, aes(x=month, y=tn_ug, color=lakename))+
scale_color_manual(values = c("Peter Lake" = "navy", 
                               "Paul Lake" = "darkorange2"))+
  #^^ manual selecting of colors for each lake
  geom_boxplot(aes(x=factor(month, levels=1:12, labels=month.abb)))+
  labs(x=element_blank(), y="total N (µg)", #eliminating x labels
       title="Monthly N levels in water of Peter and Paul Lakes")+
    theme(legend.position = "none") #eliminating legend

plot_grid(temp_plot, tp_plot, tn_plot, ncol = 1, nrow = 3) #cowplot for display
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
niwot_ridge%>%
  filter(functionalGroup == "Needles")%>%
  ggplot(aes(x=collectDate, y=dryMass, color=nlcdClass))+
  scale_color_manual(values=c("evergreenForest"="forestgreen",
                               "grasslandHerbaceous"="turquoise3",
                               "shrubScrub"="violetred2"))+
  geom_point(alpha=0.5)+
  scale_x_date(date_breaks = "3 months", date_labels = "%m%Y")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(x="date collected", y="dry mass", title="Dry mass of different kinds of plant \nneedle litter organized by date collected")


#7
ggplot(subset(niwot_ridge, nlcdClass =="evergreenForest" | 
                nlcdClass == "grasslandHerbaceous" | 
                nlcdClass == "shurbScrub"),
         aes(x=collectDate, y=dryMass))+
  geom_point(alpha=0.5)+
  scale_x_date(date_breaks = "3 months", date_labels = "%m%Y")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_grid(nlcdClass ~ dryMass)+
  labs(x="date collected", y="dry mass", title="Dry mass of different kinds of 
       plant \nneedle litter organized by date collected")
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: